---
title: "IntBio_WL2_2023"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Make a practice IPM with WL2 2023 Data

## Libraries
```{r}
library(tidyverse)
library(broom)
library(broom.mixed)
library(magrittr)
library(lmerTest)
```

## Load the data
```{r}
wl2_size_2023 <- read_csv("data/WL2-2023_Size_Combined.csv") %>% select(survey_date, Genotype, parent.pop, height.cm, long.leaf.cm)
unique(wl2_size_2023$parent.pop) #all 23 pops!
wl2_mort_2023 <- read_csv("data/WL2_Mortality_2023.csv") %>% 
  select(block, BedLoc, Genotype, parent.pop=pop, mf:death.date) %>% 
  mutate(parent.pop= str_replace(parent.pop, "Y08", "YO8")) %>% 
  mutate(parent.pop= str_replace(parent.pop, "Y04", "YO4")) %>% 
mutate(Genotype= str_replace(Genotype, "Y08", "YO8")) %>% 
  mutate(Genotype= str_replace(Genotype, "Y04", "YO4")) 
unique(wl2_mort_2023$parent.pop) #all 23 pops!

#wl2_size_2023 %>% filter(Genotype=="YO4_2_2") #I_18_C
#wl2_size_2023 %>% filter(Genotype=="YO8_8_4") #B_58_A

#wl2_mort_2023 %>% filter(Genotype=="YO4_2_2")
#wl2_mort_2023 %>% filter(BedLoc=="I_18_C")
#wl2_mort_2023 %>% filter(BedLoc=="B_58_A")
```

## Merge
```{r}
wl2_all_2023_OLD <- left_join(wl2_size_2023, wl2_mort_2023) #42 more rows, why? duplicate rows somewhere 
wl2_all_2023_OLD %>% group_by(Genotype, survey_date) %>% summarize(n=n()) %>% arrange(desc(n)) 
##Duplicates: CC_1_2, CC_9_6, IH_4_5, SC_5_6, SQ2_6_2
wl2_size_2023 %>% filter(Genotype=="CC_1_2") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="CC_9_6") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="IH_4_5") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="SC_5_6") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="SQ2_6_2") #duplicated on certain dates  

#anti_join(wl2_size_2023, wl2_mort_2023) #8 rows in the size dataframe without death date info, probably died at planting 
#anti_join(wl2_mort_2023, wl2_size_2023)

#GET RID OF DUPLICATES:
wl2_size_2023_merge_prep <- wl2_size_2023 %>% 
  arrange(Genotype, survey_date, height.cm) %>% 
  distinct(Genotype, survey_date, .keep_all=TRUE)
wl2_size_2023_merge_prep %>% filter(Genotype=="YO4_2_2")

#Merge with corrections 
wl2_all_2023 <- left_join(wl2_size_2023_merge_prep, wl2_mort_2023) %>%
  select(survey_date:parent.pop, mf:rep,block:BedLoc, height.cm, long.leaf.cm, death.date) %>% 
  drop_na(block) #get rid of plants that were not planted and never got a block assigned. 
tail(wl2_all_2023, 10)
unique(wl2_all_2023$death.date)

```

## Add in surv and sizenext columns
```{r}
wl2_all_2023_timeprep <- wl2_all_2023 %>% 
  mutate(death.date=mdy(death.date)) %>% 
  group_by(Genotype) %>% 
  mutate(height_next = lead(height.cm, order_by = survey_date), #next timepoints height
         long.leaf_next = lead(long.leaf.cm, order_by = survey_date), #next timepoints leaf length 
         survey_date_char = as.character(survey_date), #grouping variable to compare to continuous survey date 
         date_next = lead(survey_date, order_by = survey_date), #needed for surv code and interval calc
         elapsed_days= date_next - survey_date, #interval length in days
         elapsed_weeks = as.integer(elapsed_days)/7, #interval length in weeks 
         surv=if_else(is.na(death.date), 1,
                      if_else(date_next>=death.date, 0, 1))) %>% #survival to the next size time point
  ungroup() %>%
  drop_na(height.cm, height_next, surv) #for models
         
wl2_all_2023_timeprep %>% arrange(Genotype, survey_date)

unique(wl2_all_2023_timeprep$elapsed_days) #mostly 1 or 2 weeks. 20 or 23 days from initial size measurements on campus to field measurements 

#wl2_all_2023_nest <- wl2_all_2023_timeprep %>% ungroup() %>% nest(.by=survey_date)
```

## Survival Models - Using height as the main size estimte here 

```{r}
surv_models_fixed <- tribble(
  ~name,          ~f,
  "1_intercept",              "surv ~ 1", 
  "2a_linear_size",           "surv ~ height.cm", 
  "2b_linear_pops",           "surv ~ parent.pop", 
  "3a_linear_size_date_char", "surv ~ height.cm*survey_date_char",
  "3b_linear_pops_date_char", "surv ~ parent.pop*survey_date_char",
  "3c_linear_pops_date",      "surv ~ parent.pop*survey_date",
  "3d_linear_pops_date",      "surv ~ parent.pop*survey_date",
  "3e_linear_size_pops",      "surv ~ height.cm*parent.pop",
  "4a_linear_size_pops_date_char", "surv ~ height.cm*parent.pop + height.cm*survey_date_char + parent.pop*survey_date_char",
  "4b_linear_size_pops_date", "surv ~ height.cm*parent.pop + height.cm*survey_date + parent.pop*survey_date",
  "5_quadratic_size",         "surv ~ height.cm + I(height.cm^2)", 
  "6_cubic_size",             "surv ~ height.cm + I(height.cm^2) + I(height.cm^3)",
  "7a_quadratic_size_pops",    "surv ~ height.cm*parent.pop + I(height.cm^2)",
  "8a_cubic_size_pops",        "surv ~ height.cm*parent.pop + I(height.cm^2) + I(height.cm^3)",
  "9a_quadratic_full_survey_date_char", "surv ~ height.cm*parent.pop*survey_date_char + I(height.cm^2)", 
  "9b_quadratic_full_survey_date",  "surv ~ height.cm*parent.pop*survey_date + I(height.cm^2)",
  "10a_cubic_full_survey_date_char", "surv ~ height.cm*parent.pop*survey_date_char + I(height.cm^2) + I(height.cm^3)",
  "10b_cubic_full_survey_date", "surv ~ height.cm*parent.pop*survey_date + I(height.cm^2) + I(height.cm^3)"
)


#run the models 
surv_models_fixed <- surv_models_fixed %>%
  mutate(glm = map(f, ~ glm(as.formula(.), 
                            data = wl2_all_2023_timeprep, 
                            family = "binomial")), #run the models 
         predict = map(glm, predict, type = "response"), # predicting from original data...
        # type = response is for glm models, back transforms probabilities from logit scale 
         glance = map(glm, glance)) #glance at the model results

surv_models_fixed %>% select(-f, -glm) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
surv_models_fixed %>% select(-f, -glm) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 

#Linear with size and date as a character seems best 
```

```{r}
survival.model.final <- surv_models_fixed %>% filter(name == "3a_linear_size_date_char") %>% pull(glm) %>% magrittr::extract2(1)
summary(survival.model.final)
```

### Predicted vs. Observed Survival 
Calculate predictions 
```{r}
survival_curves <- wl2_all_2023_timeprep %>%
  group_by(survey_date_char) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each date
  mutate(newdata = pmap(list(survey_date_char, minSize, maxSize), \(d, x, y) tibble(survey_date_char = d,
                                                        height.cm = seq(x, y, length.out = 101)))) %>%

 # predictions for each date
  mutate(surv = map(newdata, \(nd) predict(survival.model.final, newdata = nd, type = "response"))) %>%
  
  select(-survey_date_char) %>%

  # unnest to plot
  unnest(cols = c(newdata, surv), names_repair = "minimal") 

survival_curves
```

```{r}
wl2_all_2023_timeprep %>%
  ggplot(aes(x=height.cm, y = surv)) +
  geom_point() +
  geom_line(color = "red", data = survival_curves) +
  facet_wrap(~survey_date_char, scales = "free_x")
```

## Attempt at pop = random for Survival 
```{r}
surv.models_random <- tribble(
  ~name,          ~f,
  "1a_intercept_block",              "surv ~ 1 + (1|block)", 
  "1b_intercept_pop",              "surv ~ 1 + (1|parent.pop)", 
  "1c_intercept_pop_mf",              "surv ~ 1 + (1|parent.pop/mf)", 
  "1d_intercept__pop_mf_block",              "surv ~ 1 + (1|block) + (1|parent.pop/mf)", 
  "2a_linear_size_block",           "surv ~ height.cm + (1|block)", 
  "2b_linear_size_pops",           "surv ~ height.cm + (1|parent.pop)",   "2c_linear_size_pops_mf",           "surv ~ height.cm + (1|parent.pop/mf)",
  "2d_linear_size_pops_mf_block",           "surv ~ height.cm + (1|block) + (1|parent.pop/mf)", 
  "3a_linear_size_date_char", "surv ~ height.cm*survey_date_char + (1|block) + (1|parent.pop/mf)",
  #"3b_linear_size_date",      "surv ~ height.cm*survey_date + (1|block) + (1|parent.pop/mf)",
 #Caused by error:! (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
  "5_quadratic_size",         "surv ~ height.cm + I(height.cm^2) + (1|block) + (1|parent.pop/mf)", 
  "6_cubic_size",             "surv ~ height.cm + I(height.cm^2) + I(height.cm^3) + (1|block) + (1|parent.pop/mf)",
  "9a_quadratic_full_survey_date_char", "surv ~ height.cm*survey_date_char + I(height.cm^2) + (1|block) + (1|parent.pop/mf)", 
  #"9b_quadratic_full_survey_date",  "surv ~ height.cm*survey_date + I(height.cm^2) + (1|block) + (1|parent.pop/mf)",
 #Caused by error:! (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
  #"10a_cubic_full_survey_date_char", "surv ~ height.cm*survey_date_char + I(height.cm^2) + I(height.cm^3) + (1|block) + (1|parent.pop/mf)",
 #Caused by error:! pwrssUpdate did not converge in (maxit) iterations
  #"10b_cubic_full_survey_date", "surv ~ height.cm*survey_date + I(height.cm^2) + I(height.cm^3) + (1|block) + (1|parent.pop/mf)"
 #Caused by error:! (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
)

#run the models 
surv.models_random <- surv.models_random %>%
  mutate(glmer = map(f, ~ glmer(as.formula(.), 
                            data = wl2_all_2023_timeprep, 
                            family = "binomial")), #, #run the models 
         predict = map(glmer, predict, type = "response"), 
         glance = map(glmer, tidy),
         AIC = map(glmer, AIC),
         BIC = map(glmer, BIC)) #glance at the model results

surv.models_random %>% select(-f, -glmer, -predict) %>% unnest(glance) 
surv.models_random %>% select(-f, -glmer, -predict) %>% unnest(BIC) %>% arrange(BIC) #look at the model fitting info 
surv.models_random %>% select(-f, -glmer, -predict) %>% unnest(AIC) %>% arrange(AIC) #look at the model fitting info 

#2a_linear_size_block wins with BIC
#9a_quadratic_full_survey_date_char wins with AIC
```

```{r}
survival.model.final_random <- surv.models_random %>% filter(name == "9a_quadratic_full_survey_date_char") %>% pull(glmer) %>% magrittr::extract2(1)
summary(survival.model.final_random)
```

### Predicted vs. Observed Survival 
Calculate predictions - currently not working, can't figure out how to include block, pop/mf
```{r, eval=FALSE}
survival_curves_random <- wl2_all_2023_timeprep %>%
  group_by(survey_date_char, block, parent.pop, mf) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each date
  mutate(newdata = pmap(list(survey_date_char, block, parent.pop, mf, minSize, maxSize), \(d, x, y) tibble(survey_date_char = d,
                                                        height.cm = seq(x, y, length.out = 101)))) %>%

 # predictions for each date
  mutate(surv = map(newdata, \(nd) predict(survival.model.final_random, newdata = nd, type = "response"))) %>%
  
  select(-survey_date_char) %>%

  # unnest to plot
  unnest(cols = c(newdata, surv), names_repair = "minimal") 

survival_curves_random
```

- currently not working, can't figure out how to include block, pop/mf
```{r, eval=FALSE}
wl2_all_2023_timeprep %>%
  ggplot(aes(x=height.cm, y = surv)) +
  geom_point() +
  geom_line(color = "red", data = survival_curves) +
  facet_wrap(~survey_date_char, scales = "free_x")
```

## Growth Models
Observed Patterns
```{r}
wl2_all_2023_timeprep %>% 
  #filter(elapsed_days==7) %>% 
  ggplot(aes(x=height.cm, y=height_next)) +
  geom_point(aes(color = parent.pop), alpha=.1)  +
  scale_color_viridis_d(end = 0.9) + 
  geom_abline() +
  geom_smooth(method = "lm")
```


```{r}
growth_models_fixed <- tribble(
  ~name,          ~f,
  "1_intercept",              "height_next ~ 1", 
  "2a_linear_size",            "height_next ~ height.cm", 
  "2b_linear_pops",            "height_next ~ parent.pop",
  "3a_linear_with_weeks",      "height_next ~ elapsed_weeks",
  "3b_linear_with_survey_date_char", "height_next ~ survey_date_char",
  "3c_linear_with_survey_date", "height_next ~ survey_date",
  "4a_linear_with_weeks_pops_2way", "height_next ~ height.cm*elapsed_weeks + height.cm*parent.pop + elapsed_weeks*parent.pop",
  "4b_linear_with_survey_date_char_pops_2way", "height_next ~ height.cm*survey_date_char + height.cm*parent.pop + survey_date_char*parent.pop",
  "4c_linear_with_survey_date_pops_2way", "height_next ~ height.cm*survey_date + height.cm*parent.pop + survey_date*parent.pop",
  "5_quadratic_size",         "height_next ~ height.cm + I(height.cm^2)", 
  "6_cubic_size",             "height_next ~ height.cm + I(height.cm^2) + I(height.cm^3)",
  "7a_quadratic_full_weeks",    "height_next ~ height.cm*parent.pop*elapsed_weeks + I(height.cm^2)", 
  "7b_quadratic_full_survey_date_char",    "height_next ~ height.cm*parent.pop*survey_date_char + I(height.cm^2)", 
  "7c_quadratic_full_survey_date",    "height_next ~ height.cm*parent.pop*survey_date + I(height.cm^2)", 
  "8a_cubic_full_weeks",        "height_next ~ height.cm*parent.pop*elapsed_weeks + I(height.cm^2) + I(height.cm^3)",
  "8b_cubic_full_survey_date_char",        "height_next ~ height.cm*parent.pop*survey_date_char + I(height.cm^2) + I(height.cm^3)",
  "8c_cubic_full_survey_date",        "height_next ~ height.cm*parent.pop*survey_date + I(height.cm^2) + I(height.cm^3)"
)

#run the models 
growth_models_fixed <- growth_models_fixed %>%
  mutate(lm = map(f, ~ lm(as.formula(.), data = wl2_all_2023_timeprep)), #run the models 
         predict = map(lm, predict), # predicting from original data...
         glance = map(lm, glance)) #glance at the model results

growth_models_fixed %>% select(-f, -lm) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
growth_models_fixed %>% select(-f, -lm) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 
#quadratic with 3way with survey_date as continuous wins by BIC
#cubic with 3way with survey date as character wins by AIC ... Rsquared higher for this model too
```

```{r}
growth.model.final <- growth_models_fixed %>% filter(name == "8b_cubic_full_survey_date_char") %>% pull(lm) %>% magrittr::extract2(1)
summary(growth.model.final)
```

### Predicted vs. Observed Growth
```{r}
wl2_all_2023_timeprep %>% 
  cbind(predicted={growth_models_fixed %>% filter(name=="8b_cubic_full_survey_date_char") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=height_next, y = predicted)) +
  geom_point(alpha=.2) +
  geom_abline(color="skyblue2") +
  facet_wrap(~parent.pop, scales="free")

#looks decent for most pops
```

```{r}
wl2_all_2023_timeprep %>% 
  cbind(predicted={growth_models_fixed %>% filter(name=="7c_quadratic_full_survey_date") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=height_next, y = predicted)) +
  geom_point(alpha=.2) +
  geom_abline(color="skyblue2") +
  facet_wrap(~parent.pop, scales="free")

#looks decent for most pops
#don't see much difference between the two models 
```

## Attempt at pop = random for Growth 
```{r}
growth_models_random <- tribble(
  ~name,          ~f,
  "1a_intercept_block",           "height_next ~ 1 + (1|block)", 
  "1b_intercept_pop",             "height_next ~ 1 + (1|parent.pop)",
  "1c_intercept_pop_mf",          "height_next ~ 1 + (1|parent.pop/mf)",
  "1d_intercept_pop_mf_block",    "height_next ~ 1 + (1|block) +  (1|parent.pop/mf)",
  "2a_linear_size_block",         "height_next ~ height.cm + (1|block)", 
  "2b_linear_size_pops",          "height_next ~ height.cm + (1|parent.pop)",
  "2c_linear_size_pops_mf",       "height_next ~ height.cm + (1|parent.pop/mf)",
  "2d_linear_size_pops_mf_block", "height_next ~ height.cm + (1|parent.pop)+ (1|block)",
  "3a_linear_with_weeks",         "height_next ~ elapsed_weeks + (1|block)",
  "3b_linear_with_survey_date_char", "height_next ~ survey_date_char + (1|block)",
  "3c_linear_with_survey_date",   "height_next ~ survey_date + (1|block)",
  "4a_linear_with_weeks_pops_2way", "height_next ~ height.cm*elapsed_weeks + (1|parent.pop)",
  "4b_linear_with_survey_date_char_pops_2way", "height_next ~ height.cm*survey_date_char + (1|parent.pop)",
  "4c_linear_with_survey_date_pops_2way", "height_next ~ height.cm*survey_date + (1|parent.pop)",
  "5_quadratic_size",         "height_next ~ height.cm + I(height.cm^2) + (1|block)", 
  "6_cubic_size",             "height_next ~ height.cm + I(height.cm^2) + I(height.cm^3) + (1|block)",
  "7a_quadratic_full_weeks",    "height_next ~ height.cm*elapsed_weeks + I(height.cm^2) + (1|block) +  (1|parent.pop/mf)", 
  "7b_quadratic_full_survey_date_char", "height_next ~ height.cm*survey_date_char + I(height.cm^2) + (1|block) +  (1|parent.pop/mf)", 
  "7c_quadratic_full_survey_date","height_next ~ height.cm*survey_date + I(height.cm^2) + (1|block) +  (1|parent.pop/mf)", 
  "8a_cubic_full_weeks",        "height_next ~ height.cm*elapsed_weeks + I(height.cm^2) + I(height.cm^3) + (1|block) +  (1|parent.pop/mf)",
  "8b_cubic_full_survey_date_char", "height_next ~ height.cm*survey_date_char + I(height.cm^2) + I(height.cm^3) + (1|block) +  (1|parent.pop/mf)",
  "8c_cubic_full_survey_date",    "height_next ~ height.cm*survey_date + I(height.cm^2) + I(height.cm^3) + (1|block) +  (1|parent.pop/mf)"
)

#run the models 
growth_models_random <- growth_models_random %>%
  mutate(lmer = map(f, ~ lmer(as.formula(.), data = wl2_all_2023_timeprep)), #run the models 
         predict = map(lmer, predict), # predicting from original data...
         glance = map(lmer, glance)) #glance at the model results

growth_models_random %>% select(-f, -lmer) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
growth_models_random %>% select(-f, -lmer) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 
#8b_cubic_full_survey_date_char wins by BIC and AIC
```

```{r}
growth.model.final_random <- growth_models_random %>% filter(name == "8b_cubic_full_survey_date_char") %>% pull(lmer) %>% magrittr::extract2(1)
summary(growth.model.final_random)
```

### Predicted vs. Observed Growth
```{r}
wl2_all_2023_timeprep %>% 
  cbind(predicted={growth_models_random %>% filter(name=="8b_cubic_full_survey_date_char") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=height_next, y = predicted)) +
  geom_point(alpha=.2) +
  geom_abline(color="skyblue2") +
  facet_wrap(~parent.pop, scales="free")

#looks decent for most pops
```

## P Matrix 

#### Make a dataframe to store the parameters (as in App A)
```{r}
params=data.frame(
  surv.int=NA, # Intercept from logistic regression of survival
  surv.slope1=NA, # Slope from logistic regression of survival
  surv.slope2=NA, # Quadratic slope from logistic regression of survival
  growth.int=NA, # Intercept from linear regression of growth
  growth.slope=NA, # Slope from linear regression of growth
  growth.slope2=NA, # Quadratic Slope from linear regression of growth
  growth.slope3=NA, # Cubic Slope from linear regression of growth
  growth.sd=NA # Residual sd from the linear regression of growth
)
```

#### Use broom:tidy to create a df with the coef from each model 
```{r}
surv.coefs <- surv.models_random %>% 
  filter(name == "9a_quadratic_full_survey_date_char") %>% 
  mutate(coefs = map(glmer, tidy)) %>% 
  select(name, f, coefs) %>% 
  unnest(coefs) 

params$surv.int <- surv.coefs %>% filter(term == "(Intercept)") %>% pull(estimate)
params$surv.slope1 <- surv.coefs %>% filter(term == "height.cm") %>% pull(estimate)
params$surv.slope2 <- surv.coefs %>% filter(term == "I(height.cm^2)") %>% pull(estimate)
```

```{r}
growth.coefs <- growth_models_random %>% 
  filter(name == "8b_cubic_full_survey_date_char") %>% 
  mutate(coefs = map(lmer, tidy)) %>% 
  select(name, f, coefs) %>% 
  unnest(coefs) 

params$growth.int <- growth.coefs %>% filter(term == "(Intercept)") %>% pull(estimate)
params$growth.slope <- growth.coefs %>% filter(term == "height.cm") %>% pull(estimate)
params$growth.slope2 <- growth.coefs %>% filter(term == "I(height.cm^2)") %>% pull(estimate)
params$growth.slope3 <- growth.coefs %>% filter(term == "I(height.cm^3)") %>% pull(estimate)

#Pull sigma which is the _modeled_ standard deviation of the residuals.  Merow uses observed sd of residuals.  
params$growth.sd <- growth_models_random %>% filter(name == "8b_cubic_full_survey_date_char") %>% unnest(glance) %>% pull(sigma)
```

#### Define the functions
```{r}
# 1. survival probability function
##This is inverse logit.  Provides survival probability based on size.
s.x=function(x,params) {
  u=exp(params$surv.int + params$surv.slope1*x + params$surv.slope2*x^2)
  return(u/(1+u)) 
}

# 2. growth function
## Return a probability distribution of new sizes at t+1 (xp) at a given size x.  
g.yx=function(xp,x,params) {
  dnorm(xp,mean=params$growth.int+params$growth.slope*x+params$growth.slope*x^2+params$growth.slope*x^3,sd=params$growth.sd)
}
```

#### Define the structure of the IPM
```{r}
# the sizes we are integrating over
minSize<-min(wl2_all_2023_timeprep$height.cm,na.rm=T) 
maxSize<-max(wl2_all_2023_timeprep$height.cm,na.rm=T) 

n=100 # dimensions of the matrix 

b=minSize+c(0:n)*(maxSize-minSize)/n # boundary points
y=0.5*(b[1:n]+b[2:(n+1)]) # mesh points
h=y[2]-y[1] # step size
```

#### Make the matrices (G, S, and P)
```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix

S=s.x(y,params=params) # survival at each size midpoint

P=G # placeholder; redefine P on the next line
for(i in 1:n) P[,i]=G[,i]*S[i] # growth/survival matrix
```

#### Plot the matrix
```{r}
P %>% as_tibble() %>%
  set_colnames(y) %>% #column names = each size mesh point 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) + #basic contour with fill determined by the growth*surv value 
  geom_contour(aes(z = value),lwd=.25) + #adds contour lines 
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) + #add 1:1 line 
  scale_fill_viridis_c(option = "plasma") + #change contour colors 
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Size and Growth") +
  coord_equal() + #make it a square plot 
  theme_bw()
```

Check for eviction
Comparison of predicted survival probability to observed values to test for unintentional eviction. The black line shows the fitted survival model. The red dot shows the column sums of the growth/survival matrix
```{r}
plot(y,s.x(y,params), #fitted survival model 
     xlab="Size",type="l",
        ylab="Survival Probability",lwd=12)
     points(y,apply(P,2,sum),col="red",lwd=3,cex=.1,pch=19) # column sums 
```

