---
title: "IntBio_WL2_2023"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Make a practice IPM with WL2 2023 Data

## Libraries
```{r}
library(tidyverse)
library(broom)
library(broom.mixed)
library(magrittr)
library(lmerTest)
```

## Load the data
```{r}
wl2_size_2023 <- read_csv("data/WL2-2023_Size_Combined.csv") %>% select(survey_date, Genotype, parent.pop, height.cm, long.leaf.cm)
unique(wl2_size_2023$parent.pop) #all 23 pops!
wl2_mort_2023 <- read_csv("data/WL2_Mortality_2023.csv") %>% 
  select(block, BedLoc, Genotype, parent.pop=pop, mf:death.date) %>% 
  mutate(parent.pop= str_replace(parent.pop, "Y08", "YO8")) %>% 
  mutate(parent.pop= str_replace(parent.pop, "Y04", "YO4")) %>% 
mutate(Genotype= str_replace(Genotype, "Y08", "YO8")) %>% 
  mutate(Genotype= str_replace(Genotype, "Y04", "YO4")) 
unique(wl2_mort_2023$parent.pop) #all 23 pops!

#wl2_size_2023 %>% filter(Genotype=="YO4_2_2") #I_18_C
#wl2_size_2023 %>% filter(Genotype=="YO8_8_4") #B_58_A

#wl2_mort_2023 %>% filter(Genotype=="YO4_2_2")
#wl2_mort_2023 %>% filter(BedLoc=="I_18_C")
#wl2_mort_2023 %>% filter(BedLoc=="B_58_A")
```

## Merge
```{r}
wl2_all_2023_OLD <- left_join(wl2_size_2023, wl2_mort_2023) #42 more rows, why? duplicate rows somewhere 
wl2_all_2023_OLD %>% group_by(Genotype, survey_date) %>% summarize(n=n()) %>% arrange(desc(n)) 
##Duplicates: CC_1_2, CC_9_6, IH_4_5, SC_5_6, SQ2_6_2
wl2_size_2023 %>% filter(Genotype=="CC_1_2") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="CC_9_6") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="IH_4_5") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="SC_5_6") #duplicated on certain dates  
wl2_size_2023 %>% filter(Genotype=="SQ2_6_2") #duplicated on certain dates  

#anti_join(wl2_size_2023, wl2_mort_2023) #8 rows in the size dataframe without death date info, probably died at planting 
#anti_join(wl2_mort_2023, wl2_size_2023)

#GET RID OF DUPLICATES:
wl2_size_2023_merge_prep <- wl2_size_2023 %>% 
  arrange(Genotype, survey_date, height.cm) %>% 
  distinct(Genotype, survey_date, .keep_all=TRUE)
wl2_size_2023_merge_prep %>% filter(Genotype=="YO4_2_2")

#Merge with corrections 
wl2_all_2023 <- left_join(wl2_size_2023_merge_prep, wl2_mort_2023) %>%
  select(survey_date:parent.pop, mf:rep,block:BedLoc, height.cm, long.leaf.cm, death.date) %>% 
  drop_na(block) #get rid of plants that were not planted and never got a block assigned. 
tail(wl2_all_2023, 10)
unique(wl2_all_2023$death.date)

```

## Add in surv and sizenext columns
```{r}
wl2_all_2023_timeprep <- wl2_all_2023 %>% 
  mutate(death.date=mdy(death.date)) %>% 
  group_by(Genotype) %>% 
  mutate(height_next = lead(height.cm, order_by = survey_date), #next timepoints height
         long.leaf_next = lead(long.leaf.cm, order_by = survey_date), #next timepoints leaf length 
         survey_date_char = as.character(survey_date), #grouping variable to compare to continuous survey date 
         date_next = lead(survey_date, order_by = survey_date), #needed for surv code and interval calc
         elapsed_days= date_next - survey_date, #interval length in days
         elapsed_weeks = as.integer(elapsed_days)/7, #interval length in weeks 
         surv=if_else(is.na(death.date), 1,
                      if_else(date_next>=death.date, 0, 1))) %>% #survival to the next size time point
  ungroup() %>%
  drop_na(height.cm, height_next, surv) #for models
         
wl2_all_2023_timeprep %>% arrange(Genotype, survey_date)

unique(wl2_all_2023_timeprep$elapsed_days) #mostly 1 or 2 weeks. 20 or 23 days from initial size measurements on campus to field measurements 

#wl2_all_2023_nest <- wl2_all_2023_timeprep %>% ungroup() %>% nest(.by=survey_date)
```

## Survival Models - Using height as the main size estimte here 

```{r}
surv_models_fixed <- tribble(
  ~name,          ~f,
  "1_intercept",              "surv ~ 1", 
  "2a_linear_size",           "surv ~ height.cm", 
  "2b_linear_pops",           "surv ~ parent.pop", 
  "3a_linear_size_date_char", "surv ~ height.cm*survey_date_char",
  "3b_linear_pops_date_char", "surv ~ parent.pop*survey_date_char",
  "3c_linear_pops_date",      "surv ~ parent.pop*survey_date",
  "3d_linear_pops_date",      "surv ~ parent.pop*survey_date",
  "3e_linear_size_pops",      "surv ~ height.cm*parent.pop",
  "4a_linear_size_pops_date_char", "surv ~ height.cm*parent.pop + height.cm*survey_date_char + parent.pop*survey_date_char",
  "4b_linear_size_pops_date", "surv ~ height.cm*parent.pop + height.cm*survey_date + parent.pop*survey_date",
  "5_quadratic_size",         "surv ~ height.cm + I(height.cm^2)", 
  "6_cubic_size",             "surv ~ height.cm + I(height.cm^2) + I(height.cm^3)",
  "7a_quadratic_size_pops",    "surv ~ height.cm*parent.pop + I(height.cm^2)",
  "8a_cubic_size_pops",        "surv ~ height.cm*parent.pop + I(height.cm^2) + I(height.cm^3)",
  "9a_quadratic_full_survey_date_char", "surv ~ height.cm*parent.pop*survey_date_char + I(height.cm^2)", 
  "9b_quadratic_full_survey_date",  "surv ~ height.cm*parent.pop*survey_date + I(height.cm^2)",
  "10a_cubic_full_survey_date_char", "surv ~ height.cm*parent.pop*survey_date_char + I(height.cm^2) + I(height.cm^3)",
  "10b_cubic_full_survey_date", "surv ~ height.cm*parent.pop*survey_date + I(height.cm^2) + I(height.cm^3)"
)


#run the models 
surv_models_fixed <- surv_models_fixed %>%
  mutate(glm = map(f, ~ glm(as.formula(.), 
                            data = wl2_all_2023_timeprep, 
                            family = "binomial")), #run the models 
         predict = map(glm, predict, type = "response"), # predicting from original data...
        # type = response is for glm models, back transforms probabilities from logit scale 
         glance = map(glm, glance)) #glance at the model results

surv_models_fixed %>% select(-f, -glm) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
surv_models_fixed %>% select(-f, -glm) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 

#Linear with size and date as a character seems best 
```

```{r}
survival.model.final <- surv_models_fixed %>% filter(name == "3a_linear_size_date_char") %>% pull(glm) %>% magrittr::extract2(1)
summary(survival.model.final)
```

### Predicted vs. Observed Survival 
Calculate predictions 
```{r}
survival_curves <- wl2_all_2023_timeprep %>%
  group_by(survey_date_char) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each date
  mutate(newdata = pmap(list(survey_date_char, minSize, maxSize), \(d, x, y) tibble(survey_date_char = d,
                                                        height.cm = seq(x, y, length.out = 101)))) %>%

 # predictions for each date
  mutate(surv = map(newdata, \(nd) predict(survival.model.final, newdata = nd, type = "response"))) %>%
  
  select(-survey_date_char) %>%

  # unnest to plot
  unnest(cols = c(newdata, surv), names_repair = "minimal") 

survival_curves
```

```{r}
wl2_all_2023_timeprep %>%
  ggplot(aes(x=height.cm, y = surv)) +
  geom_point() +
  geom_line(color = "red", data = survival_curves) +
  facet_wrap(~survey_date_char, scales = "free_x")
```

## Attempt at pop = random
```{r}
surv.models_random <- tribble(
  ~name,          ~f,
  "1a_intercept_block",              "surv ~ 1 + (1|block)", 
  "1b_intercept_pop",              "surv ~ 1 + (1|parent.pop)", 
  "1c_intercept_pop_mf",              "surv ~ 1 + (1|parent.pop/mf)", 
  "1d_intercept__pop_mf_block",              "surv ~ 1 + (1|block) + (1|parent.pop/mf)", 
  "2a_linear_size_block",           "surv ~ height.cm + (1|block)", 
  "2b_linear_size_pops",           "surv ~ height.cm + (1|parent.pop)",   "2c_linear_size_pops_mf",           "surv ~ height.cm + (1|parent.pop/mf)",
  "2d_linear_size_pops_mf_block",           "surv ~ height.cm + (1|block) + (1|parent.pop/mf)", 
  "3a_linear_size_date_char", "surv ~ height.cm*survey_date_char + (1|block) + (1|parent.pop/mf)",
 # "3b_linear_size_date",      "surv ~ height.cm*survey_date + (1|block) + (1|parent.pop/mf)",
 #Caused by error:! (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
  "5_quadratic_size",         "surv ~ height.cm + I(height.cm^2) + (1|block) + (1|parent.pop/mf)", 
  "6_cubic_size",             "surv ~ height.cm + I(height.cm^2) + I(height.cm^3) + (1|block) + (1|parent.pop/mf)",
  "9a_quadratic_full_survey_date_char", "surv ~ height.cm*survey_date_char + I(height.cm^2) + (1|block) + (1|parent.pop/mf)", 
  #"9b_quadratic_full_survey_date",  "surv ~ height.cm*survey_date + I(height.cm^2) + (1|block) + (1|parent.pop/mf)",
 #Caused by error:! (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
  #"10a_cubic_full_survey_date_char", "surv ~ height.cm*survey_date_char + I(height.cm^2) + I(height.cm^3) + (1|block) + (1|parent.pop/mf)",
 #Caused by error:! pwrssUpdate did not converge in (maxit) iterations
  #"10b_cubic_full_survey_date", "surv ~ height.cm*survey_date + I(height.cm^2) + I(height.cm^3) + (1|block) + (1|parent.pop/mf)"
 #Caused by error:! (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
)

#run the models 
surv.models_random <- surv.models_random %>%
  mutate(glmer = map(f, ~ glmer(as.formula(.), 
                            data = wl2_all_2023_timeprep, 
                            family = "binomial")), #, #run the models 
         predict = map(glmer, predict, type = "response"), 
         glance = map(glmer, tidy),
         AIC = map(glmer, AIC),
         BIC = map(glmer, BIC)) #glance at the model results

surv.models_random %>% select(-f, -glmer, -predict) %>% unnest(glance) 
surv.models_random %>% select(-f, -glmer, -predict) %>% unnest(BIC) %>% arrange(BIC) #look at the model fitting info 
surv.models_random %>% select(-f, -glmer, -predict) %>% unnest(AIC) %>% arrange(AIC) #look at the model fitting info 

#2a_linear_size_block wins with BIC
#9a_quadratic_full_survey_date_char wins with AIC
```

```{r}
survival.model.final_random <- surv.models_random %>% filter(name == "9a_quadratic_full_survey_date_char") %>% pull(glmer) %>% magrittr::extract2(1)
summary(survival.model.final_random)
```

### Predicted vs. Observed Survival 
Calculate predictions - currently not working, can't figure out how to include block, pop/mf
```{r, eval=FALSE}
survival_curves_random <- wl2_all_2023_timeprep %>%
  group_by(survey_date_char, block, parent.pop, mf) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each date
  mutate(newdata = pmap(list(survey_date_char, block, parent.pop, mf, minSize, maxSize), \(d, x, y) tibble(survey_date_char = d,
                                                        height.cm = seq(x, y, length.out = 101)))) %>%

 # predictions for each date
  mutate(surv = map(newdata, \(nd) predict(survival.model.final_random, newdata = nd, type = "response"))) %>%
  
  select(-survey_date_char) %>%

  # unnest to plot
  unnest(cols = c(newdata, surv), names_repair = "minimal") 

survival_curves_random
```

- currently not working, can't figure out how to include block, pop/mf
```{r, eval=FALSE}
wl2_all_2023_timeprep %>%
  ggplot(aes(x=height.cm, y = surv)) +
  geom_point() +
  geom_line(color = "red", data = survival_curves) +
  facet_wrap(~survey_date_char, scales = "free_x")
```

## Growth Models
Observed Patterns
```{r}
wl2_all_2023_timeprep %>% 
  #filter(elapsed_days==7) %>% 
  ggplot(aes(x=height.cm, y=height_next)) +
  geom_point(aes(color = parent.pop), alpha=.1)  +
  scale_color_viridis_d(end = 0.9) + 
  geom_abline() +
  geom_smooth(method = "lm")
```


```{r}
growth_models_fixed <- tribble(
  ~name,          ~f,
  "1_intercept",              "height_next ~ 1", 
  "2a_linear_size",            "height_next ~ height.cm", 
  "2b_linear_pops",            "height_next ~ parent.pop",
  "3a_linear_with_weeks",      "height_next ~ elapsed_weeks",
  "3b_linear_with_survey_date_char", "height_next ~ survey_date_char",
  "3c_linear_with_survey_date", "height_next ~ survey_date",
  "4a_linear_with_weeks_pops_2way", "height_next ~ height.cm*elapsed_weeks + height.cm*parent.pop + elapsed_weeks*parent.pop",
  "4b_linear_with_survey_date_char_pops_2way", "height_next ~ height.cm*survey_date_char + height.cm*parent.pop + survey_date_char*parent.pop",
  "4c_linear_with_survey_date_pops_2way", "height_next ~ height.cm*survey_date + height.cm*parent.pop + survey_date*parent.pop",
  "5_quadratic_size",         "height_next ~ height.cm + I(height.cm^2)", 
  "6_cubic_size",             "height_next ~ height.cm + I(height.cm^2) + I(height.cm^3)",
  "7a_quadratic_full_weeks",    "height_next ~ height.cm*parent.pop*elapsed_weeks + I(height.cm^2)", 
  "7b_quadratic_full_survey_date_char",    "height_next ~ height.cm*parent.pop*survey_date_char + I(height.cm^2)", 
  "7c_quadratic_full_survey_date",    "height_next ~ height.cm*parent.pop*survey_date + I(height.cm^2)", 
  "8a_cubic_full_weeks",        "height_next ~ height.cm*parent.pop*elapsed_weeks + I(height.cm^2) + I(height.cm^3)",
  "8b_cubic_full_survey_date_char",        "height_next ~ height.cm*parent.pop*survey_date_char + I(height.cm^2) + I(height.cm^3)",
  "8c_cubic_full_survey_date",        "height_next ~ height.cm*parent.pop*survey_date + I(height.cm^2) + I(height.cm^3)"
)

#run the models 
growth_models_fixed <- growth_models_fixed %>%
  mutate(lm = map(f, ~ lm(as.formula(.), data = wl2_all_2023_timeprep)), #run the models 
         predict = map(lm, predict), # predicting from original data...
         glance = map(lm, glance)) #glance at the model results

growth_models_fixed %>% select(-f, -lm) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
growth_models_fixed %>% select(-f, -lm) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 
#quadratic with 3way with survey_date as continuous wins by BIC
#cubic with 3way with survey date as character wins by AIC ... Rsquared higher for this model too
```

```{r}
growth.model.final <- growth_models_fixed %>% filter(name == "8b_cubic_full_survey_date_char") %>% pull(lm) %>% magrittr::extract2(1)
summary(growth.model.final)
```

### Predicted vs. Observed Growth
```{r}
wl2_all_2023_timeprep %>% 
  drop_na(height.cm, height_next) %>% 
  cbind(predicted={growth_models_fixed %>% filter(name=="8b_cubic_full_survey_date_char") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=height_next, y = predicted)) +
  geom_point(alpha=.2) +
  geom_abline(color="skyblue2") +
  facet_wrap(~parent.pop, scales="free")

#looks decent for most pops
```

```{r}
wl2_all_2023_timeprep %>% 
  drop_na(height.cm, height_next) %>% 
  cbind(predicted={growth_models_fixed %>% filter(name=="7c_quadratic_full_survey_date") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=height_next, y = predicted)) +
  geom_point(alpha=.2) +
  geom_abline(color="skyblue2") +
  facet_wrap(~parent.pop, scales="free")

#looks decent for most pops
#don't see much difference between the two models 
```

## P Matrix 


