---
title: "IPM_from_PopSim"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# IPM from Population Simulation

\*Simulations done in "Population_Sim_Trial.Rmd" file (currently using AlphaSimR)

**Goal:** Compare IPM created from just phenotypes from the simulation to an IPM created from phenotypes + genotype scores.

## Libraries

```{r}
library(tidyverse)
library(broom)
library(broom.mixed)
library(magrittr)
library(lmerTest)
conflicted::conflicts_prefer(lmerTest::lmer)
conflicted::conflicts_prefer(dplyr::filter)
```

## Load the data
```{r}
simpop <- read_csv("../output/PopSim_Phenos-Genos_v1.csv")
names(simpop)
#logit_pheno = scaled probability/phenotype
#_geno = genotype score for that trait 
#size = height calculated from simulated Weibull parameters (alpha, beta, k, and delta)
####Note: we don't have a genotype score for this since it's not explicitly estimated by the pop sim...
#elapsed_weeks = interval between weeks - should be all be 1 in this data
```

### Add size next and weekly surv
```{r}
unique(simpop$established) #NA = did not germinate, 1 = established (survived first 3 weeks in field); 0 = did not establish
unique(simpop$y1surv) #NA = did not establish, 1 = survived to week 12; 0 = did not survive 
simpop %>% select(Indiv_ID, germinated, established, y1surv, k_pheno, week, size) %>% filter(Indiv_ID<5)

simpop_timeprep <- simpop %>%
  select(-alpha_pheno, -alpha_geno, -beta_pheno, -beta_geno, -k_pheno, -k_geno, -delta_pheno, -delta_geno) %>% 
  group_by(Indiv_ID) %>% 
  mutate(size = if_else(germinated==0, NA, #if a plant did not germinate it cannot have a size at any time
                        size), #in popsim code already have week 1 size = beta (min size)
         size_next = lead(size, order_by = week), #next time point's size; *see note about not having a genotype score for this...
         surv=if_else(germinated==0, NA, #surv = NA if you did not germ 
                        if_else(week==1, 1, #if you did germinate, assume you survived to week 1 (equivalent to pre-transplant size time point)
                                if_else(established==0, 0, #surv = 0 after week 1 if you did not establish 
                                        if_else(week<5, 1, #if you did establish, your surv is equal to 1 for the first 3 weeks post-transplant 
                                                y1surv) #your survival for weeks 5-12 is then equal to your y1surv
                                        )))) %>% #note we only have establishment and y1surv survival from the pop sim --> little weekly variation 
  #also geno_score is for the pheno in logit scale... does that matter?
  ungroup() %>%
  mutate(weeks = as.numeric(week - 1), #Weeks since week 1 (equivalent to pre-transplant or initial size) 
         size = if_else(surv==0, NA, size) #no size if no surv 
         ) %>%   
  drop_na(surv) #for models

simpop_timeprep %>% filter(Indiv_ID<10) %>% select(Indiv_ID, week, weeks, germinated, established, y1surv, size, size_next, surv) #look at data for a few indivs to see if above worked
simpop %>% filter(Indiv_ID==4) #was removed from the time prep dataset since it did not germinate --> no size or surv 

simpop_timeprep_size <- simpop_timeprep %>% drop_na(size, size_next) #for size models 
```

### Scaling and transformations
```{r, eval=FALSE}
simpop_timeprep %>%  #skewed
  ggplot(aes(x=size)) +
  geom_histogram()

simpop_timeprep_scaled <- simpop_timeprep %>% 
  mutate(logSize=log(size))  #transform height to be more normal - only for surv models 
#this doesn't work because if you take NAs for size out then you only have surv = 1 ...
```

## Survival Models

### Phenos Only

#### Predicted vs. Observed Survival

### Phenos + Genos

#### Predicted vs. Observed Survival

## Growth Models

### Observed Patterns

### Phenos Only

#### Predicted vs. Observed Growth

### Phenos + Genos

#### Predicted vs. Observed Growth

## P Matrix

### Make a dataframe to store the parameters

### Use broom:tidy to create a df with the coef from each model

### Define the functions

### Define the structure of the IPM

### Make the matrices (G, S, and P)

### Plot the matrix

### Check for eviction
